Iterative Reduction: Parsing and Interpreting Code Without Trees

### 2025-01-29 Paper/paper.tex
Update paper

We can formalize the algorithm given in the introduction as follows: we start with an alphabet of tokens $\Sigma$.
Define $\oS=\Sigma\times\bbN$, and we assume the existence of a so-called \textit{initial $\beta$-reducer}:
$\hb\colon \Sigma\times\Sigma \longto \Sigma$, which is defined by the implementer of the language being interpreted, and is the function which
tells us how to combine tokens together.
We extend it to a \textit{derived $\beta$-reducer} which is a function $\beta\colon\oS^*\longto\oS^*$ as follows.
\yoni{plz explicitly say that we use $a_b$ to denote $<a,b>$.}
Suppose $\xi=\sigma^1_{p_1}\cdots\sigma^n_{p_n}$, then $\beta(\xi)$ is defined recursively as follows:
\begin{enumerate}
    \item if $n=1$, then $\beta(\xi)=\sigma^n_0$,
    \item if $p_1\geq p_2$, then $\beta(\xi)=\hb(\sigma^1,\sigma^2)_{p_2}\sigma^3_{p_3}\cdots\sigma^n_{p_n}$,
    \item otherwise $\beta(\xi)=\sigma^1_{p_1}\,\beta(\sigma^2_{p_2}\cdots\sigma^n_{p_n})$.
\end{enumerate}

@@

@@

Here we abstract away the value of each term; the value of the new term given by the initial $\beta$-reducer will be determined by an output function.
Our alphabet of tokens $\Sigma$, instead of having a token for every number, will just have tokens for types (number, operator, etc.).
We also have another set $\values$, the \textit{universe of values}.
\yoni{plz provide an example for what this universe typically looks like.}

@@

The initial $\beta$-reducer is a function
\[ \hb \colon \Sigma\times\Sigma \partial \Sigma\times(\bbZ\times\bbZ\to\bbZ)\times(\values\times\values\partial\values) \]

\yoni{plz intuitively describe what it returns.}

@@

    Define $\Sigma=\set{\NUM,\OP,\LPAREN,\RPAREN,\OP\NUM,\RPAREN\NUM}$.
    Our initial $\beta$-reducer is defined as follows:
    \begin{itemize}
        \item $\hb(\NUM,\OP)=(\OP\NUM,\snd,(v,f\mapsto(v,f))$
\yoni{This is a bit strange. It suggests that $\values\times \values \subseteq \values$, which is impossible. We need to properly define the relevant set here, and the domain and range of this function.}
        \item $\hb(\OP\NUM,\NUM)=(\NUM,\snd,((v,f),u\mapsto f(v,u)))$
        \item $\hb(\OP\NUM,\OP\NUM)=(\OP\NUM,\snd,((v,f),(u,g)\mapsto (f(v,u),g))$
        \item $\hb(\NUM,\RPAREN)=(\RPAREN\NUM,\snd,(v,\_\mapsto v))$
        \item $\hb(\OP\NUM,\RPAREN\NUM)=(\RPAREN\NUM,\snd,((v,f),u\mapsto f(v,u)))$
        \item $\hb(\LPAREN,\RPAREN\NUM)=(\NUM,\fst,(\_,v\mapsto v))$
    \end{itemize}
\yoni{Here would be a good place to show what is saved compared to the previous approach, maybe there are less lines or something?}
    In the first point $(v,f\mapsto(v,f))$ means that the function to compute the value of the new token is one which takes two values $v,f$ and outputs a new value which is the pair $(v,f)$.

    So taking our previous example of \ttt{2*(1+3)+4}: we first must convert it into tokens acceptable for $\beta$-reduction (via lexing).
    In this example allocating initial priorities is not so simple: the priorities depend not on the type of the token in $\Sigma$, but the token in the stream given to the lexer.
    So we will assume the lexer allocates priorities similar to the previous example in the previous section.
    \yoni{better to use ref instead of "previous example".}
    This issue will be resolved in the next section.
    So the stream is converted to
    \[ \NUM_\infty(2)\ \OP_2(*)\ \LPAREN_\infty\ \NUM_\infty(1)\ \OP_1(+)\ \NUM_\infty(3)\ \RPAREN_0\ \OP_1(+)\ \NUM_\infty(4) \]
    We no longer need boxes to distinguish between tokens because the individual tokens and so we will no longer use them.
    \yoni{"and so" sounds like it misses something before it.}

@@

The simplest class in our hierarchy will be $\beta$-reducible languages, the class of which we will denote by \BR.
Define a \BR-schema to be a tuple $B=(\Sigma,\Gamma,\square,\blacksquare,\bullet,s,\hb)$ where
\begin{enumerate}
    \item $\Sigma$ is a finite alphabet of terminals,
    \item $\Gamma$ is a finite alphabet of non-terminals (generally $\Sigma$ and $\Gamma$ are disjoint, but we don't require this),
    \item $\square,\blacksquare$ are symbols not in $\Sigma,\Gamma$,
    \item $\bullet$ is a symbol (which may be in $\Sigma,\Gamma$) or the empty word,
    \item $s$ is a state, which is a map $\Sigma\to\Gamma$,
    \item $\hb$ is an initial $\beta$-reducer, which is a partial function $\hb\colon\Gamma\times(\Gamma\cup\set{\epsilon,\blacksquare})\partial\Gamma\cup\set{\epsilon,\bullet}$.
\end{enumerate}
Given a \BR-schema, we define a pre-$\beta$-reducer to be a function $\beta'\colon(\Gamma\cup\set\bullet)^*\longto(\Gamma\cup\set\bullet)^*$ as follows: given a string $\xi\in(\Gamma\cup\set\bullet)^*$ we define $\beta'(\xi)$ as follows:
\begin{enumerate}
    \item If $\xi=\bullet$, then $\beta'(\xi)=\xi$
    \item If $\xi=\sigma\in\Gamma$, then $\beta'(\xi)=\hb(\sigma,\blacksquare)$ if it is defined, and $\sigma$ otherwise,
    \item If $\xi=\sigma_1\sigma_2\xi'$, then
        \begin{enumerate}
            \item $\beta'(\xi)=\hb(\sigma_1,\sigma_2)\xi'$ if it is defined, otherwise
            \item $\beta'(\xi)=\hb(\sigma_1,\epsilon)\sigma_2\xi'$ if it is defined, otherwise
            \item $\beta'(\xi)=\sigma_1\beta'(\sigma_2\xi')$.
        \end{enumerate}
\end{enumerate}
And we define a $\beta$-reducer to be a function $\beta\colon(\Gamma\cup\set\bullet)^*\longto(\Gamma\cup\set\bullet)^*$ as follows: given $\xi\in(\Gamma\cup\set\bullet)^*$ we define $\beta(\xi)$ as follows:
\begin{enumerate}
    \item if $\xi=\epsilon$ and $\hb(\square,\blacksquare)=\tau$ is defined then $\beta(\xi)=\tau$,
    \item if $\xi=\sigma\xi'$ and $\hb(\square,\sigma)=\tau$ is defined then $\beta(\xi)=\tau\xi'$,
    \item otherwise $\beta(\xi)=\beta'(\xi)$.
\end{enumerate}
Then we define $\beta^*(\xi)$ to be the fixed point of successive applications of $\beta$ to $\xi$, i.e. $\beta^*(\xi)=\beta^n(\xi)$ where $\beta^n(\xi)=\beta^{n+1}(\xi)$.
For a word $w=\sigma_1\cdots\sigma_n=\in\Sigma^*$, we define $s(w)=s(\sigma_1)\cdots s(\sigma_n)\in\Gamma^*$.
The language \textit{generated} (or \textit{accepted}) by a \BR-schema $B$ is defined to be all the words $w\in\Sigma^*$ such that $\beta^*(s(w))=\bullet$.

@@

For example, $\set{a^nb^n}[n\geq1]$ is in \BR.
We define $B=(\Sigma,\Gamma,s,\hb)$ as follows:
\begin{enumerate}
    \item $\Sigma=\set{a,b}$,
    \item $\Gamma=\set{S_a,S_b,S_{ab},S_{aab}}$,
    \item $s(a)=S_a$ and $s(b)=S_b$,
    \item $\hb$ is defined by the rules:
        \[ S_a,S_b\to S_{ab},\quad S_a,S_{ab}\to S_{aab},\quad S_{aab},S_b=S_{ab},\quad S_{ab},\blacksquare\to\bullet \]
\end{enumerate}
Indeed, a word of the form $a\cdots ab\cdots b$ will be mapped to $S_a\cdots S_aS_b\cdots S_b$, which will then be reduced to $S_a\cdots S_aS_{ab}S_b\cdots S_b\to S_a\cdots S_{aab}S_b\cdots S_b\to S_a\cdots S_{ab}\cdots S_b$, and so on until $S_{ab}$, which is then reduced
to $\bullet$.

@@

This is (we believe) stronger than \BR: $\set{ww^r}[w\in\set{a,b}^+]$ is \MBR.
We construct a $\BR(5)$-schema as follows:
\[ B = \parens{\set{a,b},\set{A,B,A',B'},\square,\blacksquare,\bullet,[a\to A,b\to B],\set{\hb_0,\hb_1,\hb_2,\hb_3,\hb_4}} \]
Where we have the reduction rules:
\begin{align*}
    \hb_0\colon\quad& \square,A\to A',1\\
        & \square,B\to B',3\\
    \hb_1\colon\quad& A,\square\to A',2\\
    \hb_2\colon\quad& A',A'\to\bullet,2\\
        &A',\epsilon\to\epsilon,2\\
        &A',\square\to A',0\\
    \hb_3\colon\quad& B,\square\to B',2\\
    \hb_4\colon\quad& B',B'\to\bullet,2\\
        &B',\epsilon\to\epsilon,2\\
        &B',\square\to B',0\\
\end{align*}
Intuitively, $\hb_0$ detects the first character and ``tags'' it, then $\hb_1$ or $\hb_3$ ``tag'' the end character if it is the same as the beginning character.
$\hb_2$ and $\hb_4$ ``clean up'' the tags: if the two tagged characters are next to each other, then we can try and accept the word.
Otherwise delete them both.

@@

But \MBR{} is not contained in \CFL{}: $\set{a^nb^nc^n}[n\geq1]\in\MBR$.
Indeed, we can construct an \MBR-schema as follows:
\[ B = \parens{\set{a,b,c},\set{A,B,C,X},[a\to A,b\to B,c\to C],\set{\hb_0,\hb_1,\hb_2,\hb_3}} \]
And we have the reduction rules:
\begin{align*}
    \hb_0\colon\quad& B,A\to X,0\\
        &C,B\to X,0\\
        &C,\square\to C,1\\
    \hb_1\colon\quad& A,\epsilon\to\epsilon,2\\
    \hb_2\colon\quad& B,\epsilon\to\epsilon,3\\
    \hb_3\colon\quad& C,\square\to\bullet,3\\
        &C,\epsilon\to\epsilon,1
\end{align*}
Intuitively, $\hb_0$ checks that $A$ comes before $B$ which comes before $C$ (as otherwise it places in an $X$, which is never reduced).
Then $\hb_1$ matches a single $A$, then calls $\hb_2$ to match a single $B$, which calls $\hb_3$ to match a single $C$.

>>>>

We can formalize the algorithm given in the introduction as follows: we start with an alphabet of tokens $\Sigma$.
Define $\oS=\Sigma\times\bbN$, and we assume the existence of a so-called \textit{initial $\beta$-reducer}:
$\hb\colon \Sigma\times\Sigma \longto \Sigma$, which is defined by the implementer of the language being interpreted, and is the function which
tells us how to combine tokens together.
We extend it to a \textit{derived $\beta$-reducer} which is a function $\beta\colon\oS^*\longto\oS^*$ as follows.
Suppose $\xi=\sigma^1_{p_1}\cdots\sigma^n_{p_n}$, where $\sigma_n$ denotes the pair $(\sigma,n)$, a token and a priority.
Then $\beta(\xi)$ is defined recursively as follows:
\begin{enumerate}
    \item if $n=1$, then $\beta(\xi)=\sigma^n_0$,
    \item if $p_1\geq p_2$, then $\beta(\xi)=\hb(\sigma^1,\sigma^2)_{p_2}\sigma^3_{p_3}\cdots\sigma^n_{p_n}$,
    \item otherwise $\beta(\xi)=\sigma^1_{p_1}\,\beta(\sigma^2_{p_2}\cdots\sigma^n_{p_n})$.
\end{enumerate}

@@

\begin{example} \label{example:partial}

@@

Here we abstract away the value of each term; the value of the new term given by the initial $\beta$-reducer will be determined by an output function.
Our alphabet of tokens $\Sigma$, instead of having a token for every number, will just have tokens for types (number, operator, etc.).
We also have another set $\values$, the \textit{universe of values}.
This could be, for example, a set consisting of all numbers, functions $\bb R\times\bb R\longto\bb R$, tuples of numbers, etc.

@@

The initial $\beta$-reducer is a function
\[ \hb \colon \Sigma\times\Sigma \partial \Sigma\times(\bbZ\times\bbZ\to\bbZ)\times(\values\times\values\partial\values) \]
That is, the initial $\beta$-reducer takes two types from $\Sigma$ and returns the following:
\begin{enumerate}
    \item a new type,
    \item a function to determine the priority of the new type given the priorities of the original two types,
    \item and a function to determine the value of the new type given the values of the original two types.
\end{enumerate}

@@

    Define $\Sigma=\set{\NUM,\OP,\LPAREN,\RPAREN,\OP\NUM,\RPAREN\NUM}$.
    And we define our universe of values to consist of
    \begin{enumerate}
        \item numbers ($\bb R$),
        \item binary operators (functions $\bb R\times\bb R\to\bb R$),
        \item pairs consisting of a number and a binary operator.
    \end{enumerate}

    Our initial $\beta$-reducer is defined as follows:
    \begin{itemize}
        \item $\hb(\NUM,\OP)=(\OP\NUM,\snd,(v,f\mapsto(v,f))$ (here $v$ is a number, $f$ a binary operator, and the result is a pair),
\yoni{This is a bit strange. It suggests that $\values\times \values \subseteq \values$, which is impossible. We need to properly define the relevant set here, and the domain and range of this function.}
            \ari{In general this is not an issue, you could for example start with a set $S_0=S$, and define $S_{n+1}=S_n\cup(S_n\times S_n)$, then $\values=\bigcup_nS_n$ has this property.
            I'm not saying this just to be technical, this is sort of how I imagined $\values$ being defined.
            In any case, I provided a simpler definition for $\values$ above.}
        \item $\hb(\OP\NUM,\NUM)=(\NUM,\snd,((v,f),u\mapsto f(v,u)))$
        \item $\hb(\OP\NUM,\OP\NUM)=(\OP\NUM,\snd,((v,f),(u,g)\mapsto (f(v,u),g))$
        \item $\hb(\NUM,\RPAREN)=(\RPAREN\NUM,\snd,(v,\_\mapsto v))$
        \item $\hb(\OP\NUM,\RPAREN\NUM)=(\RPAREN\NUM,\snd,((v,f),u\mapsto f(v,u)))$
        \item $\hb(\LPAREN,\RPAREN\NUM)=(\NUM,\fst,(\_,v\mapsto v))$
    \end{itemize}
\yoni{Here would be a good place to show what is saved compared to the previous approach, maybe there are less lines or something?}

    So taking our previous example of \ttt{2*(1+3)+4}: we first must convert it into tokens acceptable for $\beta$-reduction (via lexing).
    In this example allocating initial priorities is not so simple: the priorities depend not on the type of the token in $\Sigma$, but the token in the stream given to the lexer.
    So we will assume the lexer allocates priorities similar to example \ref{example:partial}.
    This issue will be resolved in the next section.
    So the stream is converted to
    \[ \NUM_\infty(2)\ \OP_2(*)\ \LPAREN_\infty\ \NUM_\infty(1)\ \OP_1(+)\ \NUM_\infty(3)\ \RPAREN_0\ \OP_1(+)\ \NUM_\infty(4) \]
    The boxes are removed from around the tokens because they were an aesthetic quirk we used for readability.
    Since each token is now easily distinguishable from other tokens, the boxes are no longer necessary.

@@

The simplest class in our hierarchy will be $\beta$-reducible languages, the class of which we will denote by \BR.
Define a \BR-schema to be a tuple $B=(\Sigma,\Gamma,\square,\blacksquare,\bullet,s,\hb)$ where
\begin{enumerate}
    \item $\Sigma$ is a finite alphabet of terminals,
    \item $\Gamma$ is a finite alphabet of non-terminals (we don't require $\Sigma,\Gamma$ to be disjoint),
    \item $\square,\blacksquare$ are symbols not in $\Sigma,\Gamma$,
    \item $\bullet$ is a symbol (which may be in $\Sigma,\Gamma$) or the empty word,
    \item $s$ is a state, which is a map $\Sigma\to\Gamma$,
    \item $\hb$ is an initial $\beta$-reducer, which is a partial function $\hb\colon\Gamma\times(\Gamma\cup\set{\epsilon,\blacksquare})\partial\Gamma\cup\set{\epsilon,\bullet}$.
\end{enumerate}
Given a \BR-schema, we define a pre-$\beta$-reducer to be a function $\beta'\colon(\Gamma\cup\set\bullet)^*\longto(\Gamma\cup\set\bullet)^*$ as follows: given a string $\xi\in(\Gamma\cup\set\bullet)^*$ we define $\beta'(\xi)$ as follows:
\begin{enumerate}
    \item If $\xi=\bullet$, then $\beta'(\xi)=\xi$
    \item If $\xi=\sigma\in\Gamma$, then $\beta'(\xi)=\hb(\sigma,\blacksquare)$ if it is defined, and $\sigma$ otherwise,
    \item If $\xi=\sigma_1\sigma_2\xi'$, then
        \begin{enumerate}
            \item $\beta'(\xi)=\hb(\sigma_1,\sigma_2)\xi'$ if it is defined, otherwise
            \item $\beta'(\xi)=\hb(\sigma_1,\epsilon)\sigma_2\xi'$ if it is defined, otherwise
            \item $\beta'(\xi)=\sigma_1\beta'(\sigma_2\xi')$.
        \end{enumerate}
\end{enumerate}
And we define a $\beta$-reducer to be a function $\beta\colon(\Gamma\cup\set\bullet)^*\longto(\Gamma\cup\set\bullet)^*$ as follows: given $\xi\in(\Gamma\cup\set\bullet)^*$ we define $\beta(\xi)$ as follows:
\begin{enumerate}
    \item if $\xi=\epsilon$ and $\hb(\square,\blacksquare)=\tau$ is defined then $\beta(\xi)=\tau$,
    \item if $\xi=\sigma\xi'$ and $\hb(\square,\sigma)=\tau$ is defined then $\beta(\xi)=\tau\xi'$,
    \item otherwise $\beta(\xi)=\beta'(\xi)$.
\end{enumerate}
Then we define $\beta^*(\xi)$ to be the fixed point of successive applications of $\beta$ to $\xi$, i.e. $\beta^*(\xi)=\beta^n(\xi)$ where $\beta^n(\xi)=\beta^{n+1}(\xi)$.
For a word $w=\sigma_1\cdots\sigma_n=\in\Sigma^*$, we define $s(w)=s(\sigma_1)\cdots s(\sigma_n)\in\Gamma^*$.
The language \textit{generated} (or \textit{accepted}) by a \BR-schema $B$ is defined to be all the words $w\in\Sigma^*$ such that $\beta^*(s(w))=\bullet$.

@@

For example, $\set{a^nb^n}[n\geq1]$ is in \BR.
We define $B=(\Sigma,\Gamma,\square,\blacksquare,\bullet,s,\hb)$ as follows:
\begin{enumerate}
    \item $\Sigma=\set{a,b}$,
    \item $\Gamma=\set{S_a,S_b,S_{ab},S_{aab}}$,
    \item $\square,\blacksquare,\bullet$ are all distinct elements not in $\Sigma,\Gamma$,
    \item $s(a)=S_a$ and $s(b)=S_b$,
    \item $\hb$ is defined by the rules:
        \[ S_a,S_b\to S_{ab},\quad S_a,S_{ab}\to S_{aab},\quad S_{aab},S_b=S_{ab},\quad S_{ab},\blacksquare\to\bullet \]
\end{enumerate}
Indeed, a word of the form $a\cdots ab\cdots b$ will be mapped to $S_a\cdots S_aS_b\cdots S_b$, which will then be reduced to $S_a\cdots S_aS_{ab}S_b\cdots S_b\to S_a\cdots S_{aab}S_b\cdots S_b\to S_a\cdots S_{ab}\cdots S_b$, and so on until $S_{ab}$, which is then reduced
to $\bullet$.

@@

This is (we believe) stronger than \BR: $\set{ww^r}[w\in\set{a,b}^+]$ is \MBR.
We construct a $\BR(5)$-schema as follows:
\[ B = \parens{\set{a,b},\set{A,B,A',B'},\square,\blacksquare,\bullet,[a\to A,b\to B],\set{\hb_0,\hb_1,\hb_2,\hb_3,\hb_4}} \]
Where we have the reduction rules:
\begin{align*}
    \hb_0\colon\quad& \square,A\to A',1\\
        & \square,B\to B',3\\
    \hb_1\colon\quad& A,\blacksquare\to A',2\\
    \hb_2\colon\quad& A',A'\to\bullet,2\\
        &A',\epsilon\to\epsilon,2\\
        &A',\blacksquare\to A',0\\
    \hb_3\colon\quad& B,\blacksquare\to B',2\\
    \hb_4\colon\quad& B',B'\to\bullet,2\\
        &B',\epsilon\to\epsilon,2\\
        &B',\blacksquare\to B',0\\
\end{align*}
Intuitively, $\hb_0$ detects the first character and ``tags'' it, then $\hb_1$ or $\hb_3$ ``tag'' the end character if it is the same as the beginning character.
$\hb_2$ and $\hb_4$ ``clean up'' the tags: if the two tagged characters are next to each other, then we can try and accept the word.
Otherwise delete them both.

@@

But \MBR{} is not contained in \CFL{}: $\set{a^nb^nc^n}[n\geq1]\in\MBR$.
Indeed, we can construct an \MBR-schema as follows:
\[ B = \parens{\set{a,b,c},\set{A,B,C,X},[a\to A,b\to B,c\to C],\set{\hb_0,\hb_1,\hb_2,\hb_3}} \]
And we have the reduction rules:
\begin{align*}
    \hb_0\colon\quad& B,A\to X,0\\
        &C,B\to X,0\\
        &C,\blacksquare\to C,1\\
    \hb_1\colon\quad& A,\epsilon\to\epsilon,2\\
    \hb_2\colon\quad& B,\epsilon\to\epsilon,3\\
    \hb_3\colon\quad& C,\blacksquare\to\bullet,3\\
        &C,\epsilon\to\epsilon,1
\end{align*}
Intuitively, $\hb_0$ checks that $A$ comes before $B$ which comes before $C$ (as otherwise it places in an $X$, which is never reduced).
Then $\hb_1$ matches a single $A$, then calls $\hb_2$ to match a single $B$, which calls $\hb_3$ to match a single $C$.
